Transfer learning for \'echelle spectroscopy with \texttt{blas\'e}

### 2022-04-01 paper/paper1/ms.tex
more prose on doppler imaging, RV jitter, and more

\subsection{What about minibatches?}

Minibatches offer a natural solution to the line-by-line computational bottleneck while also providing optimization performance improvements. Minibatches act as a form of regularization, the principal source of stochasticity in the Stochastic Gradient Descent algorithm, which tends to have better convergence than full-batch Gradient Descent \citep{2016arXiv160904747R}. The Machine Learning/Artificial Intelligence (ML/AI) community employs minibatches to handle similar situations, in which the dataset volume exceeds the available GPU RAM. Our configuration departs slightly from the typical ML/AI situation since their models are usually ``small'' relative to their datasets. Here our model is ``big'' relative to our dataset.

Here we choose to assemble and evaluate only a portion of the matrix at a time, in minibatches. The choice to evaluate only a portion of lines at a time would mean the model is inaccurately evaluated at all wavelength pixels. Instead, we choose to evaluate all lines, but only on a subset $N_{\mathrm{batch}}$ of the total pixels $N_s$, so that the model can eventually converge to exact at those points. All lines are allowed to update at each glimpse of a minibatch, but many lines with cores far from minibatch pixels will provide only weak information about how the loss scalar changes for their parameters. The value of $N_{\mathrm{batch}}$ should be set just below the threshold at which the computation runs out of memory. At present, this threshold is determined experimentally. The indices of the wavelength points can either be set to be contiguous in blocks, or random with possibly large gaps between indices, or else with some other scheme tailored to the most informative wavelength regions. We experimented with several techniques and settled on random sampling with replacement, meaning that some pixels get revisited more often than others, with some minibatches tailored only to the line cores.

@@

Some lines, such as Hydrogen, Sodium, potassium, neutral metal lines, and others have extremely broad line wings, approaching larger than the $\sim6000$ pixels we allocate for the sparse implementation. These special lines should be handled separately from the weak lines, both from a computational performance perspective and an accuracy perspective.

Extremely broad lines will exhibit truncation effects if the sparse window is small compared to the line wing size. The truncation effects will look like tophat functions severing the asymptotic wings, imbuing artificial step function kinks in the emulated spectrum. We can afford to increase the sparse window on a few, say $N_{broad}\sim20$ of the broadest spectral lines. We then construct and evaluate the entire dense matrix for those lines: $\sim 330\;000 \times 20$. The number of FLOPS in each category scales as about 6 Million for the 20 broad lines versus about 36 Million for the sea of about 6500 narrow lines, depending on the exact choices for wing cuts and number of lines.

@@

where $\odot$ is the element-wise product (\emph{a.k.a} Hadamard product), $\mathcal{S}$ is the sigmoid function, and we have introduced three new tunable parameters for each of the $j$ broad lines; $\lambda_t$ is the truncation wavelength, $b$ is a scale parameter for how slowly or how rapidly in wavelength-space the transition from non-Lorentzian proceeds, and $a$ is a possibly negative stretch parameter that controls whether the line wing is sub- or super- Lorentzian.

@@

>>>>

\subsection{Minibatches and Stochastic Gradient Descent}

Currently each training epoch ``sees'' the entire dataset, a setup dubbed ``full-batch'' gradient descent.  An alternative scheme allow training with only a portion of the entire dataset at a time in ``minibatches''.  The massive data volumes in modern Neural Network applications cannot fit into the GPU memory, and so minibatches are a necessity.  Our meager $1$ MB dataset can easily fit into the GPU memory, but our model can be large if we have a large number of pixels or lines or both.  So while minibatches may not be required due to data size limitations they may be useful for particularly large models.  Minibatches also act as a form of regularization, the principal source of stochasticity in the Stochastic Gradient Descent algorithm, which tends to have better convergence than full-batch Gradient Descent \citep{2016arXiv160904747R}.

We experimented with minibatches by assembling and evaluate only a portion of the dense $\bar{\bm{F}}$ matrix at a time, in minibatches. The choice to evaluate only a portion of lines at a time would mean the model is inaccurately evaluated at all wavelength pixels. Instead, we choose to evaluate all lines, but only on a random subset $N_{\mathrm{batch}}$ of the total pixels $N_s$, so that the model can eventually converge to exact at those points. All lines are allowed to update at each glimpse of a minibatch, but many lines with cores far from minibatch pixels will provide only weak information about how the loss scalar changes for their parameters. The value of $N_{\mathrm{batch}}$ should be set just below the threshold at which the computation runs out of memory. At present, this threshold is determined experimentally. The indices of the wavelength points can either be set to be contiguous in blocks (duplicating the sparse matrix approach), or random with possibly large gaps between indices, or else with some other scheme tailored to the most informative wavelength regions. We experimented with several techniques and settled on random sampling with replacement, meaning that some pixels get revisited more often than others, with some minibatches tailored only to the line cores.  Overall minibatches as implemented above performed worse than the sparse implementation, with both lower accuracy and slower computation time.

@@

Some lines, such as Hydrogen, sodium, potassium, and others have extremely broad line wings, approaching larger than the $\sim6000$ pixels we allocate for the sparse implementation. These special lines should be handled separately from the weak lines, both from a computational performance perspective and an accuracy perspective.

Extremely broad lines will exhibit truncation effects if the sparse window is small compared to the line wing size. The truncation effects will look like tophat functions severing the asymptotic wings, imbuing artificial step function kinks in the emulated spectrum. We can afford to increase the sparse window on a few, say $N_{broad}\sim20$ of the broadest spectral lines. We then construct and evaluate the entire dense matrix for those lines: $\sim 330\;000 \times 20$. The number of FLOPS in each category scales as about 6 Million for the 20 broad and dense lines versus about 36 Million for the sea of about 6500 narrow and sparse lines, depending on the exact choices for wing cuts and number of lines.

@@

where $\odot$ is again the element-wise product (\emph{a.k.a} Hadamard product), $\mathcal{S}$ is the sigmoid function, and we have introduced three new tunable parameters for each of the $j$ broad lines; $\lambda_t$ is the truncation wavelength, $b$ is a scale parameter for how slowly or how rapidly in wavelength-space the transition from non-Lorentzian proceeds, and $a$ is a possibly negative stretch parameter that controls whether the line wing is sub- or super- Lorentzian.

@@

\subsection{Wavelength dependent limb darkening}
Currently the extrinsic model step possesses up to four parameters: the $v\sin{i}$ and $RV$, and 2 optional parameters for limb-darkening.  These four parameters may adequately parameterize a star with a uniform stellar disk.  Extremely high signal-to-noise-ratio spectra of rapidly rotating stars may require additional flexibility.  The limb darkening is generally wavelength dependent, and so a pan-chomatic spectrum may require a different limb-darkening from the blue end to the red end.  The limb darkening may instead depend on physical properties of the spectral line formation, such as physical depth of formation, and so the extent of limb-darkening may jump haphazardly from line-to-line-to-line, rather than as a predictably smooth function across wavelength.  \texttt{blas\'e} could be built to handle such a seemingly pathological scenario by adding a vector of limb darkening parameters, one for each line.  One would have to regularize the fits with some typical limb darkening and a heuristic penalty for departures from this mean.

\subsection{Doppler Imaging}
The fixed $v\sin{i}$ approximation breaks down for stars with large-scale surface features.  Doppler imaging \authorcomment1{xx, yy, zz} attempts to reverse engineer the surface map from the extent to which observed line profiles depart from a pristine rotational broadening kernel.  This reverse engineering step suffers from a vast number of geometrical degeneracies, but still provides useful constraints on stellar surfaces.  Following LugerXX, we emphasize a distinction between A) longitudinally symmetric surface features, and B) longitudinally asymmetric surface features.  Most radial velocity practitioners think about the latter, since longitudinally asymmetric surface features imbue skewness to the line profiles, causing radial velocity perturbations easily detectable in radial velocity time series.  These confound exoplanet searches.

Longitudinally symmetric surface features to not change as the star rotates on its axis.  The existence of these features change the \emph{kurtosis} of the spectral line.  For example a hypothetical non-emitting (black) polar starspot exhibits a deficit in flux at the line core, resulting in less zero-velocity flux than its homogeneous counterpart.  A dark zonal band results in equal-sized bites out of the red- and blue- sides of the line.

It is much easier to reverse-engineer longitudinally asymmetric features than longitudinally symmetric ones.  The latter requires exact knowledge of the underlying spectral template.  Isolated deep spectral lines constitute the only practical scenario where exact knowledge can be claimed.  Isolated spectral lines may be scarce or absent for M-dwarfs and brown dwarfs where lines blend together ostensibly in an inseparable way, counfounding Doppler imaging.

\texttt{blas\'e} offer a new approach to Doppler imaging that may overcome these historical limitations.  In \texttt{blas\'e} can...

\subsection{RV jitter and microturbulence}

Already \texttt{blas\'e} is equipped to fit every single line with its own systemic $RV$, by tuning the line center position $\lambda_\mathrm{c}$ at each training epoch.  There exists both empirical evidence and some theoretical motivation that $RV$ jitter varies from line-to-line \authorcomment1{Xavier Dumusque 2018}.  Importantly, the extent of this line-by-line RV jitter could be predicted in part by the depth of line formation.  A future extreme-precision-RV (EPRV) version of \texttt{blase} could leverage this information in a few ways.  As an example, the depth-of-formation for all or a subset of lines could be obtained and associated with each spectral line.  A regulariztion scaling term could be introduced to allow the line positions to vary, but only in proportion to their depth of formation.
